---
title: "Data Manipulation & ggplot2 Training Session"
author: "Greg Cicconetti"
date: '2022-08-15'
output: html_document
editor_options: 
  chunk_output_type: console
---
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Data Manipulation & ggplot2 Training Session}
%\VignetteDepends{tidyverse, figuRes2, latex2exp, scales, gridExtra}

# Load packages

```{r}
require(tidyverse)
require(gridExtra)
require(scales)
require(figuRes2)
require(latex2exp)
```

# Getting started with graphics with the computer screen in mind

## Warm-up with some comparison with Base R:

### Histogram, boxplot, scatterplot in base R

Base R is great for getting a quick view of your data:

```{r}
x <- rnorm(100)
hist(x)
boxplot(x)
plot(x)
```

## Histogram, boxplot, scatterplot in ggplot

But ggplot2 offers aesthetic and programming advantages.

```{r}
my.df <- data.frame(x=x)
ggplot(data=my.df, aes(x=x)) + geom_histogram()
ggplot(data=my.df, aes(x=factor(""), y=x)) + geom_boxplot()
ggplot(data=my.df, aes(x=1:100, y=x))+geom_point()
```


# Getting aquainted with a dummy demography data set

This is dummy data set we will use to create some graphics.

```{r}
# dframe=read.csv("demog.csv")
data(demog.data)
```

## Explore data

```{r}
head(demog.data)
```

### Getting aquainted with REGION column

This a column holding info on Region.

```{r}
table(demog.data$REGION)
data.frame(table(demog.data$REGION))
```

# Bar charts 

## Suppose we want a simple univariate barchart

```{r}
ggplot(data=demog.data, aes(x=REGION)) + geom_bar()
```

## Swap X and Y Axes with coord_flip

```{r}
ggplot(data=demog.data, aes(x=REGION)) + geom_bar() + coord_flip()
```

## Assignment of ggplot object to variable

```{r}
p <- ggplot(data=demog.data, aes(x=REGION)) + geom_bar() + coord_flip()
# print or simply send p to command line
print(p)
```

## Manipulate labels and tick marks

### Add labels

```{r}
p + labs(title="Barchart of Regions", y = "Count", x="Region", subtitle="My subtitle", caption="My caption")
```

### Manipulate tickmarks

We can control the tick marks using the breaks option. It is good practice to use breaks which extend up to or beyond the planned limits of the graph.

```{r}
p + scale_y_continuous(breaks=seq(0,6000, 500), limits= c(0, 5000))

# Note that tick mark at 5000 is not printed.
p + scale_y_continuous(breaks=seq(0, 4500, 500), limits= c(0, 5000))
```

### Altogether now...

```{r}
p + labs(title="Barchart of Regions", y = "Count", x="Region", subtitle="My subtitle", caption="My caption")+ scale_y_continuous(breaks=seq(0,4000, 500))
```

## Manipulating the orders factor levels are plotted

To exert control over the order of categories, we need to convert REGION from a character vector into a factor vector.

```{r}
class(demog.data)
demog.data$REGION
class(demog.data$REGION)
levels(demog.data$REGION) # <-- we need to convert this to a factor
```

### Default factor ordering upon coercion of vectors into factors

If a character vector is coerced into a factor, the default ordering is lexicographical.  

```{r}
levels(factor(demog.data$REGION)) 

# Indeed, Alphabetical ordering
all.equal(levels(factor(demog.data$REGION)),
          levels(factor(demog.data$REGION, c("Asia/Pacific",
                                         "Eastern Europe",
                                         "North America",
                                         "South America",  
                                         "Western Europe"))))
```

### Dictating the order of a factor manually

Suppose the following order is desired. E.g., we wish to see Eastern/Western Europe and North/South America to be ordered together.

```{r}
levels(factor(demog.data$REGION, c("Asia/Pacific",
                               "Eastern Europe",
                               "Western Europe",
                               "North America",
                               "South America")))
```

### Dictating order programmatically

Suppose we wish to order categories by their frequency. The base::order function returns a permutation which rearranges its first argument into ascending or descending order.

```{r}
# We really want to order these by frequency...
data.frame(original.order = levels(factor(demog.data$REGION)),
           original.frequency = data.frame(table(demog.data$REGION))$Freq,
           map=order(table(demog.data$REGION)),
           new.order=levels(factor(demog.data$REGION))[order(table(demog.data$REGION))])
```

Let's declare two new column vector, REGION2, REGION3:

```{r}
demog.data$REGION2 <- factor(demog.data$REGION, c("Asia/Pacific",
                                          "Eastern Europe",
                                          "North America",
                                          "South America",  
                                          "Western Europe")[order(table(demog.data$REGION))])

demog.data$REGION3 <- factor(demog.data$REGION, c("Asia/Pacific",
                                          "Eastern Europe",
                                          "North America",
                                          "South America",  
                                          "Western Europe")[rev(order(table(demog.data$REGION)))])
```

### Altogether now...

Here we use gridExtra::grid.arrange to juxtapose ggplot objects.

```{r}
p1 <- ggplot(data=demog.data, aes(x=REGION)) + geom_bar() + coord_flip()
p2 <- ggplot(data=demog.data, aes(x=REGION2)) + geom_bar() + coord_flip()
p3 <- ggplot(data=demog.data, aes(x=REGION3)) + geom_bar() + coord_flip()

grid.arrange(p1 + labs(title="Default ordering - alphabetical"), 
             p2 + labs(title="Ordered by frequency"),
             p3 + labs(title="Ordered by Frequency & reversed"), ncol=1)
```

## Bar charts from summarized data

Suppose instead of beginning with subject level data, we begin with aggregated summaries.

```{r}
# Note the use of dplyr::rename to rename columns
demog.data.table.original <- data.frame(table(demog.data$REGION))
demog.data.table.renamed <- demog.data.table.original %>% rename(REGION = Var1)

demog.data.table.original
demog.data.table.renamed
```

```{r}
ggplot(data=demog.data.table.renamed, aes(x=REGION, y=Freq)) + geom_bar(stat="identity")
```

## Suppose we want relative frequency bar chart

```{r}
demog.data.table.original$prop <- demog.data.table.original$Freq/sum(demog.data.table.original$Freq)
ggplot(data=demog.data.table.original, aes(x=Var1, y=prop)) + geom_bar(stat="identity")
```

Challenge: Manipulate the limits and tick marks so this graphic's y-axis extends to 0.3 and shows tick marks at 0, 0.025, 0.05, ..., 0.30.

## A punched up version - barchart by region and gender

Items to note: 

* ggplot2::aes links the color used to fill in bars to Var2 (SEX)
* ggplot2::geom_bar takes position = "dodge", contrast this with the default: position="stack"
* ggplot2::geom_text is employed to add frequency counts
* scales::percent_format is used to add % signs on axis. The scales packages is a good package to review!
* ggplot2::scale_fill_manual is used to manipulate the colors used to fill bars
* ggplot2::theme is used to manipulate the position of the legend

```{r}
demog.data.table <- data.frame(table(demog.data$REGION, demog.data$SEX))
demog.data.table$prop <- demog.data.table$Freq/sum(demog.data.table$Freq)

p1 <- ggplot(data = demog.data.table, aes(x = Var1, y = prop, fill = Var2)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  coord_flip() + 
  geom_text(position=position_dodge(width=1), size = 5, 
             aes(y=prop+.02, label=paste(" n = ", Freq, sep = "")))+
  scale_y_continuous(limits = c(0,.3), breaks = seq(0,1,.05), labels = percent_format()) + 
  scale_fill_manual(values = c("blue", "brown")) + 
  labs(y = "Proportion", x = "Region", fill = "Gender", title="Barchart of Region and Gender")+
  theme(legend.position= "bottom")
p1
```


# Plotting parametric functions

## Base R approach

Here's base R's graphics::curve function in action. It provides a graphic and if assigned to an objects, returns a list.

```{r}
curve(expr = dnorm(x), from=-6, to=6, n=1001)
temp <- curve(expr = dnorm(x), from=-6, to=6, n=1001)
class(temp)
str(temp)
```

The list readily converts to a data.frame which is needed for a ggplot graphic.

```{r eval=F}
head(data.frame(temp))
```

## the figuRes2::gcurve function

This function is a hack of the graphics::curve function.  It suppresses graphic creation, returns a data.frame and optionally provides an additional column to the data.frame called 'category'.

```{r }
head(figuRes2::gcurve(expr = dnorm(x), from=-6, to=6, n=101, category="Standard Normal"))
```

By using multiple calls to gcurve and using dplyr::bind_rows we can quickly create graphics with multiple functions.

Note the use of color within ggplot2::aes.  

```{r }
my.df <- rbind(
  gcurve(expr = dnorm(x), from=-6, to=6, n=101, category="N(0, 1)"),
  gcurve(expr = dnorm(x, -1, 1), from=-6, to=6, n=101, category="N(-1, 1)"),
  gcurve(expr = dnorm(x, 1, 1), from=-6, to=6, n=101, category="N(1, 1)"))

ggplot(data=my.df, aes(x=x, y=y, color=category))+geom_line()
```

Here's a punched-up version:

* size = .75 is used within ggplot2::geom_line to manually control the line size
* y-label is suppressed by populating it with a blank character vector.
* the second graphics contrasts this by populated the y-labels with NULL
   * note how this increases with width of the figure
* the third graphic uses the argument breaks=NULL in ggplot2::scale_y_continuous to remove tick marks and axis labels. (This maybe helpful when comparing density curves. The y-axis may distract colleagues from appreciating that area under curves is 1 for probability density functions.)

```{r }
grid.arrange(top='Comparison of labs(y=""), labs(y=NULL), and labs(y=NULL) & scale_y_continuous(breaks=NULL)',
ggplot(data=my.df, aes(x=x, y=y, color=category))+
  geom_line(size=.75)+
  labs(x="x", y="", 
       title="Normal Density Functions", 
       color="Density")+
  theme(legend.position="bottom"),

ggplot(data=my.df, aes(x=x, y=y, color=category))+
  geom_line(size=.75)+
  labs(x="x", y=NULL, 
       title="Normal Density Functions", 
       color="Density")+
  theme(legend.position="bottom"),
ggplot(data=my.df, aes(x=x, y=y, color=category))+
  geom_line(size=.75)+
  labs(x="x", y=NULL, 
       title="Normal Density Functions", 
       color="Density")+
  theme(legend.position="bottom")+
  scale_y_continuous(breaks=NULL)
)
```



## Run through betas with a common mean

I'm looking to explore beta distributions that share a common mean.

Things to remember: 

```{r}

```

# Histograms

Let's start from scratch:

```{r }
head(demog.data)
str(demog.data)
sort(names(demog.data))
```

## Basic historgram

Here's a histogram from heights.

```{r }
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram()
summary(demog.data$HEIGHT)
```

###  Investigating the warning message

We see

Warning message:
Removed 27 rows containing non-finite values (stat_bin). 

```{r eval=T}
# note na.rm = T is needed
max(demog.data$HEIGHT, na.rm=T) - min(demog.data$HEIGHT, na.rm=T)
median(demog.data$HEIGHT)
mean(demog.data$HEIGHT)
median(demog.data$HEIGHT, na.rm=T)
mean(demog.data$HEIGHT, na.rm=T)
```

### binwidth options

```{r eval=T}
grid.arrange(top="Comparison of various binwidth values passed to geom_histogram",
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram(binwidth = .5) + ggtitle("binwidth = .5"),
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram(binwidth = 1)+ ggtitle("binwidth = 1"),
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram(binwidth = 2.5)+ ggtitle("binwidth = 2.5"),
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram(binwidth = 25)+ ggtitle("binwidth = 25"),
ncol=2)
```


## Understanding color and fill options **within the geom_historgram function**

Here we demonstrate the difference between color and fill options.  

```{r eval=T}
grid.arrange(top="Comparison of various binwidth values passed to geom_histogram when color and fill differ",
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram(binwidth = .5, color="red", fill="blue") + ggtitle("binwidth = .5"),
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram(binwidth = 1, color="red", fill="blue")+ ggtitle("binwidth = 1"),
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram(binwidth = 2.5, color="red", fill="blue")+ ggtitle("binwidth = 2.5"),
ggplot(data=demog.data, aes(x=HEIGHT)) + geom_histogram(binwidth = 25, color="red", fill="blue")+ ggtitle("binwidth = 25"),
ncol=2)
```

## Understanding alpha option for controling fill transparency

```{r eval=T}
q <- ggplot(data=demog.data, aes(x=HEIGHT)) +  ggtitle("binwidth = 2.5") 

grid.arrange(top="Comparison of various values passed to transparency argument alpha",
  q + geom_histogram(binwidth = 2.5, color="red", fill="blue", alpha=0) + labs(subtitle = "alpha = 0"),
  q + geom_histogram(binwidth = 2.5, color="red", fill="blue", alpha=.2) + labs(subtitle = "alpha = 0.2"),
  q + geom_histogram(binwidth = 2.5, color="red", fill="blue", alpha=.4) + labs(subtitle = "alpha = 0.4"),
  q + geom_histogram(binwidth = 2.5, color="red", fill="blue", alpha=.6) + labs(subtitle = "alpha = 0.6"),
  q + geom_histogram(binwidth = 2.5, color="red", fill="blue", alpha=.8) + labs(subtitle = "alpha = 0.8"),
  q + geom_histogram(binwidth = 2.5, color="red", fill="blue", alpha=1) + labs(subtitle = "alpha = 1.0")
)
```

## manipulating tick marks with breaks and plot window with limits

```{r eval=T}
 q + geom_histogram(binwidth = 2.5, color="red", fill="blue", alpha=.6) + 
  scale_x_continuous(breaks=seq(0,500,10), limits=c(100,225)) + 
  scale_y_continuous(breaks=seq(0, 3000, 100)) +
  labs(subtitle = "alpha = 0.6", caption="Scaling of axes accomplished with:\nscale_x_continuous(breaks=seq(0,500,10), limits=c(100,225))\nscale_y_continuous(breaks=seq(0, 3000, 250))")
```

## Histograms with two categories of subjects

It's important to note that these are 'stacked' histograms. It is easy for us to think these are 'overlayed' histograms.
In the first graphic, Females occupy the first factor level and we see this results in Males getting plotted first with Females stacked on top. In the second graphic, "Men" occupy the first factor level. 

```{r eval=T}
demog.data$SEX <- factor(demog.data$SEX, c("F", "M")) # Alphabetical ordering
demog.data$SEX2 <- factor(demog.data$SEX, c("M", "F")) # Preferred order
table(demog.data$SEX)
table(demog.data$SEX2)

grid.arrange(top="Comparison of plots when swapping order of factor levels",
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5),
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX2)) + 
  geom_histogram(bin=2.5))
```

I.e., transparency is not an issue here.
```{r eval=T}
grid.arrange(top="Comparison of plots when swapping order of factor levels - transparency added",
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2),
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX2)) + 
  geom_histogram(bin=2.5, alpha=.2))

```

## Histograms with facet_wrap

Note that various options serve to obscure the inherent differences in samples sizes between the two groups.

```{r eval=T}
grid.arrange(top="Comparison of facet_wrap scales options",
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2) + facet_wrap(~SEX) + ggtitle("facet_wrap default"),
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2) + facet_wrap(~SEX, scales="free_x") + ggtitle('scales="free_x"'),
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2) + facet_wrap(~SEX, scales="free_y") + ggtitle('scales="free_y"'),
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2) + facet_wrap(~SEX, scales="free") + ggtitle('scales="free"'), ncol=2)

grid.arrange(top="Comparison of facet_wrap scales options with single column of facets imposed",
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2) + facet_wrap(~SEX, ncol=1) + ggtitle("facet_wrap default, ncol=1"),
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2) + facet_wrap(~SEX, scales="free_x", ncol=1) + ggtitle('scales="free_x", ncol=1'),
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2) + facet_wrap(~SEX, scales="free_y", ncol=1) + ggtitle('scales="free_y", ncol=1'),
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX)) + 
  geom_histogram(bin=2.5, alpha=.2) + facet_wrap(~SEX, scales="free", ncol=1) + ggtitle('scales="free", ncol=1'), ncol=2)
```

## HOMEWORK!!! Go look online for solution - how do you create a relative frequency histogram??

# Boxplot

Let's build up a boxplot towards a standard.

```{r}
get.whiskers <- function(dframe) {
                bplot <- boxplot(dframe$RESPONSE ~ dframe$CATEGORY, plot = F)
                whiskers <- with(bplot, 
                                 data.frame(CATEGORY = names, 
                                            LOWER = stats[1, ], 
                                            MEDIAN = stats[3, ], 
                                            UPPER = stats[5, ],
                                            N=n))
                return(whiskers)
        }
        
whiskers <- get.whiskers(dframe = demog.data %>% select(SUBJID, TRTGRP, BMI) %>% rename(RESPONSE=BMI, CATEGORY=TRTGRP ))

demog.data %>% select(SUBJID, TRTGRP, BMI) %>% ggplot(aes(x=TRTGRP, y = BMI, shape=TRTGRP, fill=TRTGRP)) + geom_boxplot()

demog.data %>% select(SUBJID, TRTGRP, BMI) %>% ggplot(aes(x=TRTGRP, y = BMI, shape=TRTGRP, fill=TRTGRP)) +
geom_boxplot(outlier.size = 3, outlier.colour = alpha("black", 0.2))+
         stat_summary(fun = mean, geom = "point", 
                             shape = c(21, 22), 
                             size = 3, bg = "white") +
                scale_fill_manual(values = c("red", "blue")) +
                scale_x_discrete(breaks = levels(demog.data$TRTGRP), 
                                 labels = paste(levels(demog.data$TRTGRP),"\n n = ", whiskers$N)) + 
                scale_y_continuous(limits = c(0,100), breaks = seq(0,100,10)) + 
                geom_text(data = whiskers %>% rename(TRTGRP=CATEGORY) %>% mutate(x=as.numeric(as.factor(TRTGRP)) - .5), aes(x = x, y = MEDIAN, label = round(MEDIAN, 2)), 
                          size = 5) + 
                theme(legend.position= "bottom")+ labs(fill="Treatment Group")+
                labs(y = TeX(paste0("BMI kg/", "$m^{2}$")), 
                     x = "Treatment Group", 
                     fill = "Treatment Group",
                     shape = "Treatment Group",
                     title=TeX(paste0("Boxplot of BMI kg/", "$m^{2}$")),
                     caption="The median value is displayed to the left of each boxplot.")
```

# Violin plots behave very similarly to boxplots 
```{r eval=T}
ggplot(data=demog.data, aes(x=factor(""), y=HEIGHT)) + geom_violin() 
ggplot(data=demog.data, aes(x=factor(""), y=HEIGHT)) + geom_violin() + labs(x="All Patients", y="Height")
ggplot(data=demog.data, aes(x=SEX, y=HEIGHT, fill=BMI.GRP)) + geom_violin()

ggplot(data=demog.data, aes(x=SEX, y=HEIGHT, fill=SEX)) + geom_violin() + facet_wrap(~REGION)
ggplot(data=demog.data, aes(x=SEX, y=HEIGHT, fill=SEX)) + geom_violin() + facet_wrap(~REGION, nrow=1)
ggplot(data=demog.data, aes(x=SEX, y=HEIGHT, fill=SEX)) + geom_violin() + facet_wrap(BMI.GRP~REGION)
ggplot(data=demog.data, aes(x=SEX, y=HEIGHT, fill=SEX)) + geom_violin() + facet_grid(REGION~BMI.GRP)

```

## Problems: need to reorder levels of BMI.GRP

* Need to suppress the missing group
* Need to move the legend to make better use of real estate
* This chunk of code is to get the superscripts into labels

```{r eval=T}
# First get rid of missing category
demog.data2 <- subset(demog.data, BMI.GRP != "Missing")
levels(demog.data$BMI.GRP)
# Problem 1 - levels are out of order
demog.data2$BMI.GRP <- factor(demog.data2$BMI.GRP, c("<25 kg/m2", "25-<30 kg/m2", ">=30 kg/m2" , "Missing"))
levels(demog.data2$BMI.GRP)
# Better
ggplot(data=demog.data2, aes(x=SEX, y=HEIGHT, fill=SEX)) + geom_violin() + facet_grid(REGION~BMI.GRP)
# Punching it up; Note use of escape character '\n' for new line.
p <- ggplot(data=demog.data2, aes(x=SEX, y=HEIGHT, fill=SEX)) + geom_violin() + facet_grid(REGION~BMI.GRP)+
  theme(legend.position="none", strip.text=element_text(size=13)) + labs(x="Gender", y="Height", title="Violin plots of Height by Gender\nfor each BMI Group and Region")
p

demog.data2$BMI.GRP2 <- demog.data2$BMI.GRP
levels(demog.data2$BMI.GRP2) <- c(TeX(paste0("<25 kg/", "$m^2$")),
                                 TeX(paste0("25-<30 kg/", "$m^2$")),
                                 TeX(paste0("<25 kg/", "$m^2$")),
                                 TeX(paste0("$\\leq 30$", "kg/", "$m^2$")))
t1 <- expression(paste("Asia & Pacific"))
t2 <- expression(paste("Eastern Europe"))
t3 <- expression(paste("North America"))
t4 <- expression(paste("South America"))
t5 <- expression(paste("Western Europe"))
demog.data2$REGION2 <- factor(demog.data2$REGION, levels =levels(demog.data2$REGION), labels=c(t1, t2, t3, t4, t5) )

# note that Region does not throw an error here
ggplot(data=demog.data2, aes(x=SEX, y=HEIGHT, fill=SEX)) + geom_violin() + facet_grid(REGION~BMI.GRP2)+
  theme(legend.position="none", strip.text=element_text(size=13)) + labs(x="Gender", y="Height", title="Violin plots of Height by Gender\nfor each BMI Group and Region")

# note that Region throws an error here
ggplot(data=demog.data2, aes(x=SEX, y=HEIGHT, fill=SEX)) + geom_violin() + facet_grid(REGION~BMI.GRP2, labeller = label_parsed)+
  theme(legend.position="none", strip.text=element_text(size=13)) + labs(x="Gender", y="Height", title="Violin plots of Height by Gender\nfor each BMI Group and Region")


```



```{r eval=T}
# A little more work to get symbols - this is just to whet your appetite
# Redefining factor with 3 levels.  These are expressions that allow superscripts.
# look at help files for phantom - it will introduce you to displaying mathematical expressions
# 
temp1 <- expression(paste(phantom()<=25," kg/",m^2))
temp2 <- expression(paste("25-<30 kg/", m^2))
temp3 <- expression(paste(phantom() >= 30, " kg/", m^2))
demog.data2$BMI.GRP2 <- factor(demog.data2$BMI.GRP, levels=c("<25 kg/m2", "25-<30 kg/m2", ">=30 kg/m2"),labels=c(temp1, temp2, temp3))


```

# Unfortunately this step is required since 'label_parsed' is being used in the facet_grid command below

```{r eval=T}
t1 <- expression(paste("Asia/ Pacific"))
t2 <- expression(paste("Eastern Europe"))
t3 <- expression(paste("North America"))
t4 <- expression(paste("South America"))
t5 <- expression(paste("Western Europe"))
```

# So REGION2 makes use of levels and labels, 
```{r eval=T}
demog.data2$REGION2 <- factor(demog.data2$REGION, levels =levels(demog.data2$REGION), labels=c(t1, t2, t3, t4, t5) )

d <- ggplot(data=demog.data2, aes(x=SEX, y=HEIGHT, fill=SEX)) + 
  geom_violin() + 
  facet_grid(REGION2~BMI.GRP2,labeller="label_parsed")+
  labs(x="Gender", y="Height", title=expression(paste("Gender vs. Height by Region and BMI Group" )))+
  theme(legend.position="none", strip.text=element_text(size=13)) + 
  labs(x="Gender", y="Height", title="Violin plots of Height by Gender\nfor each BMI Group and Region")

d
```


# Again, geom_violin and geom_boxplot are interchangable
```{r eval=T}
ggplot(data=demog.data2, aes(x=SEX, y=HEIGHT, fill=SEX)) + 
  geom_boxplot() + 
  facet_grid(REGION2~BMI.GRP2,labeller="label_parsed")+
  labs(x="Gender", y="Height", title=expression(paste("Gender vs. Height by Region and BMI Group" )))+
  theme(legend.position="none", strip.text=element_text(size=13)) + labs(x="Gender", y="Height", title="Violin plots of Height by Gender\nfor each BMI Group and Region")

```

# Now many of the ideas we've seen will carry over to geom_density
```{r eval=T}
ggplot(data=demog.data2, aes(x=HEIGHT, fill=SEX))+geom_density()
# Now transparency really means something
ggplot(data=demog.data2, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)

ggplot(data=demog.data2, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_wrap(~REGION)
ggplot(data=demog.data2, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_wrap(~REGION, nrow=1) + theme(legend.position="bottom")
# Making use of the work we did earlier for superscripts
ggplot(data=demog.data2, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom")
```


# Continuous vs. continuous

```{r eval=T}
# explore csv file
head(demog.data)

# build table for REGION - a variable holding info on Region
table(demog.data$REGION)
```

```{r eval=T}
# Let's grab a sample
useme <- sample(x=1:nrow(demog.data), size=300)
demog.data2 <- demog.data[useme,]
# Overplotting obscures
ggplot(data=demog.data2, aes(x=REGION, y= HEIGHT)) + geom_violin() + geom_point()
# Jittering helps to see the spread
ggplot(data=demog.data2, aes(x=REGION, HEIGHT, colour=REGION)) + geom_violin() + geom_point(position="jitter")
# Violin and boxplot are interchangable
ggplot(data=demog.data2, aes(x=REGION, HEIGHT)) + geom_boxplot() + geom_point(position="jitter")
```

# Let's look at everything
```{r eval=T}
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT)) + geom_point()
# Jittering is also helpful with scatterplots
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT)) + geom_point(position="jitter")
# Transparency helps
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT)) + geom_point(position="jitter", alpha=.2)

# Back to the smaller set to see color
ggplot(data=demog.data2, aes(x=HEIGHT, y=WEIGHT, color=REGION)) + geom_point()
# Shape mapped to SEX
ggplot(data=demog.data2, aes(x=HEIGHT, y=WEIGHT, color=REGION, shape=SEX)) + geom_point()
# Let's increase the size
ggplot(data=demog.data2, aes(x=HEIGHT, y=WEIGHT, color=REGION, shape=SEX)) + geom_point(size=3)

# We can map size to a factor
ggplot(data=demog.data2, aes(x=HEIGHT, y=WEIGHT, color=REGION, shape=SEX, size=factor(BMI.GRP))) + geom_point()

```

# We can manipulate the how size is mapped

```{r eval=T}
ggplot(data=demog.data2, aes(x=HEIGHT, y=WEIGHT, color=REGION, shape=SEX, size=(factor(BMI.GRP)))) + geom_point()+  scale_size_manual(breaks=levels(demog.data2$BMI.GRP), values=c(1,2,3,10))

#We can map size to a continuous endpoint
ggplot(data=demog.data2, aes(x=HEIGHT, y=WEIGHT, color=REGION, shape=SEX, size=BMI)) + geom_point()

# Facetting works
ggplot(data=demog.data2, aes(x=HEIGHT, y=WEIGHT, color=REGION, shape=SEX, size=factor(BMI.GRP))) + geom_point()+
  scale_size_manual(breaks=levels(demog.data2$BMI.GRP), values=c(1,2,3,10)) + 
  facet_grid(SEX~BMI.GRP)+
  labs(x="Height", y="Weight", color="Region", size="BMI Group", shape="Gender")+
  theme(legend.position="bottom")+
  guides(shape="none", size="none")+
  ggtitle("Height vs. Weight by BMI Group and Gender")

```


```{r eval=T}
# We can map color to scatter 
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT, color=SEX)) + geom_point(position="jitter", alpha=.2)
# We can add contours to scatterplot
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT, color=SEX)) + geom_point(position="jitter", alpha=.2)+ geom_density2d()
# Note that to override contour color we do so within the geom
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT, color=SEX,linetype=SEX)) + geom_point(position="jitter", alpha=.2)+ geom_density2d(color="black")
# Facetting
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT, color=SEX,linetype=SEX)) + geom_point(position="jitter", alpha=.2)+ geom_density2d(color="black")+facet_grid(SEX~REGION)+
  theme(legend.position="bottom")
```


# Adding reference, OLS lines and smoothers to scatterplots

```{r}

```

# Finding Outliers
```{r eval=T}
# help(mahalanobis)
# Assign the mahalanobis distance to the dataframe
# MD is chi-sq when data is MV normal distributed.
demog.data.sub <- subset(demog.data, select=c("HEIGHT", "WEIGHT"))
demog.data$MD.HW <- mahalanobis(demog.data.sub, center=colMeans(demog.data.sub, na.rm=T), cov=cov(demog.data.sub,use="pairwise.complete.obs"))
ggplot(data=demog.data, aes(x=MD.HW))+geom_density()
#qqplot - suggest more outliers than chisq
ggplot(data=demog.data, aes(sample=MD.HW))+ stat_qq(dist=qchisq, dparam=list(df=2))
# culprits highlighted in scatterplot
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT, color=MD.HW)) + geom_point(position="jitter")+
  scale_colour_gradient(limits=c(12, 75), low="orange", high="red")
# Exaggeration
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT, color=MD.HW, size=MD.HW)) + geom_point(position="jitter")+
  scale_colour_gradient(limits=c( qchisq(df=2,.99), 75), low="orange", high="red")+
  labs(x="Height", y="Weight", size="Distance", color="Distance", title="Height vs. Weight with Mahalanobis Distance\nYellow-Red scale used for Distances beyond 99th percentile of Chisq(2)")
# Note: in practice we should split data by gender, compute MD based on Gender-specific means and covarainces
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT, color=MD.HW, size=MD.HW, shape=SEX)) + geom_point(position="jitter")+
  scale_colour_gradient(limits=c( qchisq(df=2,.99), 75), low="orange", high="red")+
  labs(x="Height", y="Weight", size="Distance", color="Distance", title="Height vs. Weight with Mahalanobis Distance\nYellow-Red scale used for Distances beyond 99th percentile of Chisq(2)")


# Faceting works again
ggplot(data=demog.data, aes(x=HEIGHT, y=WEIGHT, color=MD.HW, size=MD.HW, shape=SEX)) + geom_point(position="jitter")+
  scale_colour_gradient(limits=c( qchisq(df=2,.99), 75), low="orange", high="red")+
  labs(x="Height", y="Weight", size="Distance", color="Distance", title="Height vs. Weight with Mahalanobis Distance\nYellow-Red scale used for Distances beyond 99th percentile of Chisq(2)")+
  facet_wrap(~SEX)

```


# Add a section dedicated to theme manipulation

```{r}
# theme_bw
# Making use of the work we did earlier for superscripts
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom") + theme_bw()

# theme_light
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom") + theme_light()

# theme_dark
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom") + theme_dark()

# theme_minimal
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom") + theme_minimal()

# theme_classic
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom") + theme_classic()

# theme_void
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom") + theme_void()

# theme_test
ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom") + theme_test()
```

# How about creating a theme of our own...

I love the look of the generic theme_gray but...

* I hate that legend.position defaults to right of graphic - I like it on the bottom

```{r}
theme_gray <- function (base_size = 11, base_family = "", 
                        base_line_size = base_size/22, 
                        base_rect_size = base_size/22){

half_line <- base_size/2
t <- theme(
        line = element_line(colour = "black", size = base_line_size, 
                            linetype = 1, lineend = "butt"), 
        rect = element_rect(fill = "white", colour = "black", size = base_rect_size, linetype = 1), 
        text = element_text(family = base_family, face = "plain", 
                            colour = "black", size = base_size, lineheight = 0.9, 
                            hjust = 0.5, vjust = 0.5, angle = 0, margin = margin(), 
                            debug = FALSE), 
        axis.line = element_blank(), 
        axis.line.x = NULL, 
        axis.line.y = NULL, 
        axis.text = element_text(size = rel(0.8), colour = "grey30"), 
        axis.text.x = element_text(margin = margin(t = 0.8 * half_line/2), vjust = 1), 
        axis.text.x.top = element_text(margin = margin(b = 0.8 * half_line/2), vjust = 0), 
        axis.text.y = element_text(margin = margin(r = 0.8 * half_line/2), hjust = 1), 
        axis.text.y.right = element_text(margin = margin(l = 0.8 * half_line/2), hjust = 0), 
        axis.ticks = element_line(colour = "grey20"), 
        axis.ticks.length = unit(half_line/2, "pt"), 
        axis.ticks.length.x = NULL, 
        axis.ticks.length.x.top = NULL, 
        axis.ticks.length.x.bottom = NULL, 
        axis.ticks.length.y = NULL, 
        axis.ticks.length.y.left = NULL, 
        axis.ticks.length.y.right = NULL, 
        axis.title.x = element_text(margin = margin(t = half_line/2), vjust = 1), 
        axis.title.x.top = element_text(margin = margin(b = half_line/2), vjust = 0), 
        axis.title.y = element_text(angle = 90, margin = margin(r = half_line/2), vjust = 1),
        axis.title.y.right = element_text(angle = -90, margin = margin(l = half_line/2), vjust = 0),
        legend.background = element_rect(colour = NA), 
        legend.spacing = unit(2 * half_line, "pt"), 
        legend.spacing.x = NULL, 
        legend.spacing.y = NULL, 
        legend.margin = margin(half_line, half_line, half_line, half_line), 
        legend.key = element_rect(fill = "grey95", colour = NA), 
        legend.key.size = unit(1.2, "lines"), 
        legend.key.height = NULL, 
        legend.key.width = NULL, 
        legend.text = element_text(size = rel(0.8)), 
        legend.text.align = NULL, 
        legend.title = element_text(hjust = 0), 
        legend.title.align = NULL, 
        legend.position = "right", 
        legend.direction = NULL, 
        legend.justification = "center", 
        legend.box = NULL, 
        legend.box.margin = margin(0, 0, 0, 0, "cm"), 
        legend.box.background = element_blank(), 
        legend.box.spacing = unit(2 * half_line, "pt"), 
        panel.background = element_rect(fill = "grey92", colour = NA), 
        panel.border = element_blank(), 
        panel.grid = element_line(colour = "white"), 
        panel.grid.minor = element_line(size = rel(0.5)), 
        panel.spacing = unit(half_line, "pt"), 
        panel.spacing.x = NULL, 
        panel.spacing.y = NULL, 
        panel.ontop = FALSE, 
        strip.background = element_rect(fill = "grey85", colour = NA), 
        strip.text = element_text(colour = "grey10", size = rel(0.8), 
                                  margin = margin(0.8 * half_line, 
                                                  0.8 * half_line, 
                                                  0.8 * half_line, 
                                                  0.8 * half_line)), 
        strip.text.x = NULL, 
        strip.text.y = element_text(angle = -90), 
        strip.text.y.left = element_text(angle = 90), 
        strip.placement = "inside", 
        strip.placement.x = NULL, 
        strip.placement.y = NULL, 
        strip.switch.pad.grid = unit(half_line/2, "pt"),
        strip.switch.pad.wrap = unit(half_line/2, "pt"), 
        plot.background = element_rect(colour = "white"), 
        plot.title = element_text(size = rel(1.2), hjust = 0, vjust = 1, 
                                  margin = margin(b = half_line)), 
        plot.title.position = "panel", 
        plot.subtitle = element_text(hjust = 0, vjust = 1, margin = margin(b = half_line)), 
        plot.caption = element_text(size = rel(0.8), hjust = 1, 
                                    vjust = 1, margin = margin(t = half_line)), 
        plot.caption.position = "panel", 
        plot.tag = element_text(size = rel(1.2), hjust = 0.5, vjust = 0.5), 
        plot.tag.position = "topleft", 
        plot.margin = margin(half_line, half_line, half_line, half_line), 
        complete = TRUE)
ggplot_global$theme_all_null %+replace% t
}
```


```{r}
data(salaries, package="car")
library(ggplot2)
mytheme <- theme(plot.title=element_text(face="bold.italic",
                 size="14", color="brown"),
                 axis.title=element_text(face="bold.italic",
                    size=10, color="brown"),
                 axis.text=element_text(face="bold", size=9,
                    color="darkblue"),
                 panel.background=element_rect(fill="white",
                    color="darkblue"),
                 panel.grid.major.y=element_line(color="grey",
                    linetype=1),
                 panel.grid.minor.y=element_line(color="grey",
                    linetype=2),
                 panel.grid.minor.x=element_blank(),
                 legend.position="top")


ggplot(data=demog.data, aes(x=HEIGHT, fill=SEX))+geom_density(alpha=.3)+facet_grid(SEX~REGION) + theme(legend.position="bottom") +labs(title="Here's the title") +
        theme(plot.title=element_text(face="bold.italic",
                 size="14", color="brown"),
                 axis.title=element_text(face="bold",
                    size=10, color="brown"),
                 axis.text=element_text(face="bold", size=9,
                    color="darkblue"),
                 panel.background=element_rect(fill="white",
                    color="darkblue"),
                 panel.grid.major.y=element_line(color="grey",
                    linetype=1),
                 panel.grid.minor.y=element_line(color="grey",
                    linetype=2),
                 panel.grid.minor.x=element_blank(),
                 legend.position="top")
```


# Add a section on Color palattes

[Colorbrewer](https://colorbrewer2.org/)                
[Dealing with colors in ggplot2](https://r-graph-gallery.com/ggplot2-color.html)
[Top R Color Palettes](https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/)

```{r}
library("viridis")          
library(ggplot2)
# Gradient color
ggplot(iris, aes(Sepal.Length, Sepal.Width))+
  geom_point(aes(color = Sepal.Length)) +
  scale_color_viridis(option = "D")+
  theme_minimal() +
  theme(legend.position = "bottom")

# Discrete color. use the argument discrete = TRUE
ggplot(iris, aes(Sepal.Length, Sepal.Width))+
  geom_point(aes(color = Species)) +
  geom_smooth(aes(color = Species, fill = Species), method = "lm") + 
  scale_color_viridis(discrete = TRUE, option = "D")+
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  theme(legend.position = "bottom")

library(RColorBrewer)
display.brewer.all()

display.brewer.pal(n = 8, name = 'Dark2')
```


```{r}
library("ggplot2")
library("ggsci")
# Change area fill color. JCO palette
ggplot(iris, aes(Species, Sepal.Length)) +
  geom_boxplot(aes(fill = Species)) +
  scale_fill_jco()+
  theme_classic() +
  theme(legend.position = "top")

# Change point color and the confidence band fill color. 
# Use tron palette on dark theme
ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_point(aes(color = Species)) +
  geom_smooth(aes(color = Species, fill = Species)) + 
  scale_color_tron()+
  scale_fill_tron()+
  theme_dark() +
  theme(
    legend.position = "top",
    panel.background = element_rect(fill = "#2D2D2D"),
    legend.key = element_rect(fill = "#2D2D2D")
    )
```


```{r}
barplot(1:5, col=rainbow(5))
# Use heat.colors
barplot(1:5, col=heat.colors(5))
# Use terrain.colors
barplot(1:5, col=terrain.colors(5))
# Use topo.colors
barplot(1:5, col=topo.colors(5))
# Use cm.colors
barplot(1:5, col=cm.colors(5))
```


